---
title: "Assignment Regression models Course Project"
output: html_document
---
####Yann K / Regression models assignment due Apr,17 2016
###Executive Summary
 1. Questions from Motor Trend Magazine 
    + Are automatic transmissions better than manual transmissions in terms of consumption in miles per gallon (US)?
    + What is the quantitative difference between manual and automatic transmissions? 
 2. Key findings from the `mtcars` dataset(1974) on 32 cars models 
    + There is a significant difference between manual and automatic transmissions. 
    + Using a multivariate regression that explains 81% of the variation of `mpg` using the explanatory variables `drat`, `qsec`, `vs`, `am` and `carb` we can estimate that the automatic transmissions use in average 5.45 gallons/mile less than manual transmissions with a level of confidence of 99%.
 
###Data Exploration
```{r, echo=FALSE}
library(datasets)
library(ggplot2)
library(car)
library(grid)
data(mtcars)
# Converting am and vs variables to factors
mtcars$am<-factor(mtcars$am)
mtcars$vs<-factor(mtcars$vs)
# Auto and manual transmissions data subsets
mtcars_auto <- subset(mtcars, am==0)
mtcars_man <- subset(mtcars, am==1)

# Frequency histogram of the consumption of the sampled histograms
g <- ggplot(data=mtcars, aes(mpg, fill=am))
g <- g + geom_histogram(bins=10)
g <- g + scale_fill_discrete(name="Transmission", breaks=c("0", "1"), label=c("Automatic", "Manual"))
g <- g + geom_vline(xintercept=mean(mtcars_auto$mpg), colour="red", linetype="dashed", size=1) 
g <- g + geom_vline(xintercept=mean(mtcars_man$mpg), colour="darkgreen", linetype="dashed", size=1) 
g <- g + ggtitle("Consumption Frequency Histogram") + xlab("Consumption (Miles/Gallon US)")

# Density histogram of the consumption of the sampled automobiles
g1 <- ggplot(data=mtcars, aes(x=mpg))
g1 <- g1 + geom_histogram(aes(y=..density..), bins=10, fill="blue")
g1 <- g1 + stat_function(fun=dnorm, args=list(mean=mean(mtcars$mpg), sd=sd(mtcars$mpg)), colour="red", size=2)
g1 <- g1 + ggtitle("Consumption Density Histogram") + xlab("Consumption (Miles/Gallon US)")

#boxplot per transmission type
g2 <- ggplot(data=mtcars, aes(x=am, y=mpg, fill=am))
g2 <- g2 + geom_boxplot()
g2 <- g2 + stat_summary(fun.y=mean, geom = "point", shape=8, size=4)
g2 <- g2 + scale_fill_discrete(name="Transmission", breaks=c("0", "1"), label=c("0-Automatic", "1-Manual"))

#Plot the above 3 graphs on a 2x2 grid
grid.newpage()
pushViewport(viewport(layout=grid.layout(2,2)))
print(g, vp=viewport(layout.pos.row=1, layout.pos.col=1))
print(g1, vp=viewport(layout.pos.row=1, layout.pos.col=2))
print(g2, vp=viewport(layout.pos.row=2, layout.pos.col=1))

#T-Test to check if there is any significant difference in the consumption of automatic and manual transmissions
t.test(mtcars_auto$mpg, mtcars_man$mpg)

```
#### Conclusion from the exploratory analysis
 * The consumption of the sampled automobiles appear to be approximatively normally distributed
 * The risk of skewed data is low as there are no outliers in the sample
 * From the results of the T-test on the difference of average consumptions between automatic and manual transmissions subsets, there is a significant difference in average consumptions between the 2 types of transmissions. Let's quantity the difference with linear regression models.
 
### Simple linear regression
```{r}
fs <- lm(mpg~am, data=mtcars)
summary(fs)$coefficients
summary(fs)$r.squared
```
####Conclusion from the simple linear regression
It is estimated with a 95% confidence interval that manual transmissions are using an extra 7.2 gallons/mile. But the model explains only 36% of the variation of `mpg`. Let's use a multivariate regression model to improve the R squared.

### Multivariate linear regression
```{r}
#Multivariate linear regression using all the variables
#The p-value 0.23 for the factor am is not significant
tail(summary(lm(mpg~., data=mtcars))$coefficients, 3)
summary(lm(mpg~., data=mtcars))$r.squared
##Correlation matrix on all combinations of variables excluding the 2 factor variables (am, vs)
round(cor(subset(mtcars, select=-c(am,vs))), 2)
##Using variance inflation factor to remove highly correlated variables
round(vif(lm(mpg~., data=mtcars)), 2)
#Rerunning multivariate regression without variable disp, which got the highest vif score with 21.62
#The p-value for factor am is not significant still
tail(summary(lm(mpg~., data=subset(mtcars, select=-c(disp))))$coefficients, 3)
#After 5 iterations of removing high vif score variables (disp, cyl, hp, wt, gear)
fm<-lm(mpg~., data=subset(mtcars, select=-c(disp, cyl, hp, wt,gear)))
#The p-value for factor am is significant p-value=0.00316 < 0.01
tail(summary(fm)$coefficients, 3)
summary(fm)$r.squared
#Annex: anova between the simple linear regression and the multivariate one 
anova(fs, fm)
```
### Conclusion
 * The first tentative to run a linear regression using all the variables results on a poor fit for `am` (p-value=0.23 > 0.05)
 * This is due to the high collinearite between the explanatory variables as shown in the correlation matrix with a coefficient of correlation of 90% between `cyl` and `disp` for example.
 * We use the variance inflation factor (VIF) iteratively in 2 steps:
    + Remove the variable with the highest VIF score at each step. At the first run, we remove `disp` with a VIF score of 21.62
    + Rerun the multivariate linear regression on a subset excluding the above variables and check the p-value of factor am from the regression output. At the first run, `am` has still p-value=0.23 > 0.05
 * After 5 iterations, and the removal of `disp`, `cyl`, `hp`, `wt` and `gear`, we are able to get (p-value=0.003 < 0.05) for `am` from a multivariate regression that explains about 81% of the variation of `mpg`
 * We can conclude that there is a 5.45 gallons/mile consumption difference between automatic and manual transmissions with a level of significance of 99% as the p-value of `am` is lower than 0.01.

```{r}
#Annex: Residual plots show that they are normally distributed
par(mfrow = c(2, 2))
plot(fm)
```
